---
title: "snRNAseq_AmbRNA_Removal"
author: "Kevin Green"
date: "7/8j /2020"
output: pdf_document
---

```{r setup, cache=FALSE, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(Seurat)
library(Matrix)
library(ggplot2)
library(SoupX)
library(foreach)
library(doParallel)
library(stringr)
library(parallel)
library(scran)
library(scater)
```

# Jayadev Lab RNAseq Data Processing Pipeline

## Define your project folders, filenames, and variables of interest.
# You MUST edit this for your dataset to run properly!!!!
```{r 'Project_Information', echo=TRUE}
#Document your code:
print("This is when this code was run:")
print(Sys.time())

#Load image from snRNAseq_QC_code workspace with ss_data_split object
print("Loading thresholding_pu1_only.rdata")
load("cluster_ws_pu1_only.rdata")

#Set sample type (i.e. "nuclei", "pbmcs", "organoid" or "other")
sample_type <- "nuclei"

```
#Create list with cluster information for use with SoupX
```{r}

#Create an array for DR objects
DR_objects <- rep(NA, length(ss_data_split))

for (file in 1:length(ss_data_split)) {
  df <- as.data.frame(clusters[[file]]@reducedDims@listData$UMAP)
  df <- cbind(df, clusters[[file]]@colData@listData$clusters)
  colnames(df)[3] <- "Cluster"
  sample <- paste(names[file], "_DR", sep = "")
  DR_objects[file] <- list(sample = assign(sample, df))
  names(DR_objects)[file] <- sample
  rm(list = sample)
}

#Remove objects that are no longer needed
rm(df, clusters)
rm(list = names)

```

```{r Create SoupX objects and add cluster info}

#Check if souppath directory exists or needs to be created.
souppath <- file.path(outpath, "souppath", fsep = "/")
if (file.exists(souppath)) {
  cat("Souppath directory path works!")
  } else {
    cat("Soupath directory does not exist - creating")
    dir.create(souppath)
  }

#Create an array to hold the SoupX objects
soupx_objects <- rep(NA, length(samples))

#Create a list of SoupX objects from each data file
for (file in 1:length(samples_path)) {
  sparse_data <- Seurat::Read10X(data.dir = samples_path[[file]])
  soup_object <- SoupX::SoupChannel(tod = sparse_data,
                                    toc = ss_data_split[[file]]@assays$RNA@counts,
                                    soupRange = c(0, 10),
                                    keepDroplets = FALSE)       
  soupx_objects[file] <- list(assign(x=names[[file]],
                                       soup_object))
  names(soupx_objects)[file] <- names[[file]]
  rm(list = names[file])
}

#Remove objects that are no longer needed
rm(sparse_data, soup_object, ss_data_split)

#Sets the dimensionality reduction for all future plots
for (file in 1:length(soupx_objects)) {
  soupx_objects[[file]] <- setDR(soupx_objects[[file]],
                                 DR_objects[[file]])
}

#Sets cluster info to determined clusters in DR object
for (file in 1:length(soupx_objects)) {
  soupx_objects[[file]] <- setClusters(soupx_objects[[file]],
                                       DR_objects[[file]]$Cluster)
}
# Save soupx_objects file if needed to retest later
save.image(file = paste0("soupx_objects_ws", suffix, ".rdata"))

```

```{r Explore and set genes to be used for soup estimation}
#Explore highly expressed mRNA in soup to determine good candidates
for (file in 1:length(soupx_objects)) {
  soup_profile <- head(soupx_objects[[file]]$soupProfile[order(soupx_objects[[file]]$soupProfile$est, decreasing = TRUE), ],
       n = 100)
  
  write.csv(soup_profile, file = paste(souppath, "/",
                                       names[[file]],
                                       "_soup_profile.csv",
                                       sep = ""))
}

# This will save a marker gene plot for each sample. These plots
# can be used to determine good candidates for marker genes
for (file in 1:length(soupx_objects)) {
  plot12 <- plotMarkerDistribution(soupx_objects[[file]])

  png(paste(souppath, "/", names[[file]],
            "_soup_marker_plot", suffix, ".png", sep = ""))
    print(plot12)
  dev.off()
}

#Remove temporary objects
rm(soup_profile, plot12)

```

#This code will be used to establish gene sets that are expressed
#only in one cell type for use with estimating contamination. This
#may need to be adjusted for each dataset. If examining nuclei, use #hemoglobin genes as below
```{r Establishing gene sets for contamination estimation}

# #Establish KCN gene list to use as estimation for non-expressed
# #cells
# geneList <- c("SOX2")
# 
# #Explore KCN genes as possible marker genes
# for (file in 1:length(soupx_objects)) {
#   for(gene in 1:length(geneList)) {
#     plot13 <- plotMarkerMap(soupx_objects[[file]],
#                             geneList[[gene]])
# 
#     png(paste(souppath, "/", names[[file]], "_", geneList[[gene]],
#               "_marker_plot.png", sep = ""), 800, 800)
#       print(plot13)
#     dev.off()
#   }
# }
# 
# rm(plot13)
# 
# # #Set KCN genes to those expressed in neurons (for 2 initial
# # #samples)
# # geneList <- c("KCNC2", "KCNH1", "KCNIP4")
# 
# est_list <- rep(NA, length(soupx_objects))
# 
# #Establish which cells do not express KCN genes, and should be
# #used to estimate contamination
# for (file in 1:length(soupx_objects)) {
#   use_to_est <- estimateNonExpressingCells(soupx_objects[[file]],
#                                         nonExpressedGeneList =
#                                           list(IG = geneList))
#   est_list[file] <- list(assign(paste(names[[file]]), use_to_est))
#   names(est_list)[file] <- names[[file]]
#   rm(list = names[file])
# }
# 
# #Print a marker plot for all genes being used to
# #estimate contamination
# for (file in 1:length(soupx_objects)) {
#   plot14 <- plotMarkerMap(soupx_objects[[file]],
#                           geneSet = geneList,
#                           useToEst = use_to_est[[file]])
# 
#   png(paste(souppath, "/", names[[file]],
#             "_gene_list_marker_plot.png", sep = ""), 800, 800)
#     print(plot14)
#   dev.off()
# }
# 
# rm(plot14, use_to_est)

```

#This will create hemoglobin and mitochondrial gene lists, establish which nuclei will be used to estimate the contamination (should be all), plot these nuclei, and create counts for 
```{r Calculate contamination fraction}

hb_genes <- c("HBA1", "HBA2", "HBB")

igGenes = c('IGHA1','IGHA2','IGHG1','IGHG2','IGHG3','IGHG4','IGHD','IGHE','IGHM',
            'IGLC1','IGLC2','IGLC3','IGLC4','IGLC5','IGLC6','IGLC7',
            'IGKC')

mito_genes <- rep(NA, length(soupx_objects[[1]]$toc@Dimnames[[1]]))
for (gene in 1:length(soupx_objects[[1]]$toc@Dimnames[[1]])) {
  if (startsWith(soupx_objects[[1]]$toc@Dimnames[[1]][[gene]], "MT-") == TRUE) {
  mito_genes[[gene]] <- soupx_objects[[1]]$toc@Dimnames[[1]][[gene]]
  }
}

#Remove empty cells in mito_genes list
mito_genes <- Filter(function(x)!all(is.na(x)), mito_genes)

if (sample_type == "nuclei") {
  geneList <- mito_genes
  gene_type <- "mitochondrial genes"
} else if (sample_type == "pbmcs") {
  geneList <- igGenes
  gene_type <- "IG genes"
} else if (sample_type == "organoid") {
  geneList <- c("SOX2")
  gene_type <- "SOX2"
} else {
  geneList <- ""
}

est_list <- rep(NA, length(soupx_objects))

if (geneList != "") {
  #Establish which cells do not express mitochondrial genes, and should
  #be used to estimate contamination
  for (file in 1:length(soupx_objects)) {
    use_to_est <- estimateNonExpressingCells(soupx_objects[[file]],
                                          nonExpressedGeneList =
                                            list(IG = geneList))
    # for (i in 1:length(use_to_est)) {
    #   use_to_est[i] <- TRUE
    # }
    if (!any(use_to_est)) {
      use_to_est <- estimateNonExpressingCells(soupx_objects[[file]],
                                               nonExpressedGeneList =
                                               list(IG = geneList),
                                               maximumContamination = 2)
    }
    est_list[file] <- list(assign(paste(names[[file]]), use_to_est))
    names(est_list)[file] <- names[[file]]
    rm(list = names[file])
  }
  
  #Print a marker plot for all nuclei being used to estimate
  #contamination
  for (file in 1:length(soupx_objects)) {
    plot16 <- plotMarkerMap(soupx_objects[[file]],
                            geneSet = geneList,
                            useToEst = est_list[[file]])
  
    png(paste(souppath, "/", names[[file]],
              "_mito_genes_marker_plot", suffix, ".png", sep = ""), 800, 800)
      print(plot16)
    dev.off()
  }
  
  rm(plot16, use_to_est)
  
  #Use mitochondrial genes to estimate contamination
  for (file in 1:length(soupx_objects)) {
    soupx_objects[[file]] <-
      calculateContaminationFraction(soupx_objects[[file]],
                                     list(genes = geneList),
                                     useToEst = est_list[[file]])
    writeLines(paste("Estimated fraction of ambient RNA in", names[[file]],
                "\n using only", gene_type, "is",
                round((soupx_objects[[file]]$metaData$rho[1])*100, 1)))
  }
  
  #Estimate the contamination fraction using iterative process with
  #starting point from mitochondrial genes
  for (file in 1:length(soupx_objects)) {
    if (soupx_objects[[file]]$metaData$rho[1] > .05) {
      soupx_objects[[file]] <- 
        autoEstCont(soupx_objects[[file]],
                    priorRho = soupx_objects[[file]]$metaData$rho[1],
                    verbose = 2)
      writeLines(paste0("Estimated fraction of ambient RNA in ",
                   names[[file]],"\n using rho from non-", gene_type,
                   " expressing droplets is",
                   round((soupx_objects[[file]]$metaData$rho[1])*100, 1),
                   "%"))
      if (soupx_objects[[file]]$metaData$rho[1] < .05) {
        soupx_objects[[file]] <- 
          setContaminationFraction(soupx_objects[[file]], 0.05)
        writeLines(paste("Estimated fraction of ambient RNA in",
                         names[[file]],"is less 5%. Setting to 5%"))
      }
    }
  }
} else {
  for (file in 1:length(soupx_objects)) {
    soupx_objects[[file]] <- 
      autoEstCont(soupx_objects[[file]], verbose = 2)
    writeLines(paste("Estimated fraction of ambient RNA in", names[[file]],
                "using using default rho is",
                round((soupx_objects[[file]]$metaData$rho[1])*100, 1),
                "%"))
    if (soupx_objects[[file]]$metaData$rho[1] < .05) {
        soupx_objects[[file]] <- 
          setContaminationFraction(soupx_objects[[file]], 0.05)
        writeLines(paste("Estimated fraction of ambient RNA in",
                         names[[file]],"is less 5%. Setting to 5%"))
    }
  }
}

for (file in 1:length(soupx_objects)) {
  if (soupx_objects[[file]]$metaData$rho[1] > .2) {
    writeLines(paste("Estimated fraction of ambient RNA in", names[[file]],
                "is greater than 20%. Recalculating."))
    soupx_objects[[file]] <- autoEstCont(soupx_objects[[file]],
                                         priorRho = .2,
                                         verbose = 2)
    writeLines(paste("Estimated fraction of ambient RNA in", names[[file]],
                "\n using default rho is",
                round((soupx_objects[[file]]$metaData$rho[1])*100, 1),
                "%"))
    if (soupx_objects[[file]]$metaData$rho[1] < .05) {
        soupx_objects[[file]] <- 
          setContaminationFraction(soupx_objects[[file]], 0.05)
        writeLines(paste("Estimated fraction of ambient RNA in",
                         names[[file]], "is less than 5%. Setting to 5%"))
    }
    if (soupx_objects[[file]]$metaData$rho[1] > .2) {
        soupx_objects[[file]] <- 
          setContaminationFraction(soupx_objects[[file]], 0.2)
        writeLines(paste("Estimated fraction of ambient RNA in", 
                         names[[file]],"is greater than 20%. Setting to 20%"))
    }
  }
}
```

```{r estimates}
# Print out a list of the final estimates
for (file in 1:length(soupx_objects)) {
  writeLines(paste("The ambient RNA fraction removed from", names[[file]],
              "will be",
              round((soupx_objects[[file]]$metaData$rho[1])*100, 1),
              "%"))
}
```

```{r remove_contamination}
#Create vector to populate with adjusted count files
#post-contamination removal
soupx_out <- rep(NA, length(soupx_objects))

#Removes the ambient RNA from each sample
remove_ambient <- function(data) {
    adjustCounts(data)
}

system.time(soupx_out <- lapply(soupx_objects, remove_ambient
                                #mc.cores = cores
                                ))

```

#This code will explore the genes that were most effected by
#removal and any specific gene we want to look at. It is not
#necessary to run this portion.
```{r}
#Explore genes that were reduced to zero
for (file in 1:length(soupx_objects)) {
  pre_removal <- rowSums(soupx_objects[[file]]$toc > 0)
  post_removal <- rowSums(soupx_out[[file]] > 0)
  most_zeroed = tail(sort((pre_removal -
                             post_removal)/pre_removal),
                     n = 10)
  
  write.csv(most_zeroed, file = paste(souppath, "/",
                                      names[[file]],
                                      "_most_zeroed", suffix, ".csv",
                                      sep = ""))
}

rm(pre_removal, post_removal, most_zeroed)

#Explore the genes with the largest quantitative difference
for (file in 1:length(soupx_objects)) {
  changed_genes <- tail(sort(rowSums(soupx_objects[[file]]$toc >
                                       soupx_out[[file]])/
                                        rowSums(soupx_objects[[file]]$toc 
                                                          > 0)),
                        n = 20)
  
  write.csv(changed_genes, file = paste(souppath, "/",
                                      names[[file]],
                                      "_most_changed", suffix, ".csv",
                                      sep = ""))
}

rm(changed_genes)

#This explores specific genes
genes_to_explore <- list("S100B", "MT-CO2", "HBA1", "HBA2", "HBB")

for (gene in 1:length(genes_to_explore)) {
  for (file in 1:length(soupx_objects)) {
    plot17 <- plotChangeMap(soupx_objects[[file]], soupx_out[[file]],
                            genes_to_explore[[gene]])
    
    png(paste(souppath, "/", names[[file]], "_", genes_to_explore[[gene]],
              "_change_plot", suffix, ".png", sep = ""), 800, 800)
      print(plot17)
    dev.off()
  }
}

rm(plot17, soupx_objects)
```
  
```{r Create seurat object}
#Create Seurat objects from decontaminated data for further analysis
for (file in 1:length(soupx_out)){
  seurat_obj <- CreateSeuratObject(counts = soupx_out[[file]], 
                                   project = samples[[file]])
  assign(x=names[[file]], seurat_obj)
}

#Join the individual patient matrices into one
if (length(samples) > 1) {
  ss_data_decon <- merge(get(names[1]),get(names[2]))
  if (length(samples_path) >= 3){
    for (obj in 3:length(samples_path)){
      ss_data_decon <- merge(ss_data_decon, get(names[obj]))
      }
  }
} else {
  ss_data_decon <- get(names[1])
}

#Remove objects that are no longer needed
rm(seurat_obj, DR_objects, est_list, genes_to_explore, soupx_out)
for (file in 1:length(names)) {
  rm(list = names[file])
}

#Save workspace image if needed to revisit
save.image(file = paste0("soupX_ws", suffix, ".rdata"), compress = TRUE)

#Save the merged Seurat object for further analysis
saveRDS(ss_data_decon, file = paste0("ss_data_decon", suffix, ".rds"))
```

## Rerun thresholds for your dataset to remove low quality nuclei
```{r}
# Rerun thresholding after removal of ambient RNA counts

# ss_data_decon[["percent.mito"]] <- PercentageFeatureSet(object = ss_data_decon, pattern = "^MT-")
# 
# ss_data_decon_filtered <- subset(x = ss_data_decon, 
#                            subset = nFeature_RNA >= 350 & 
#                              nCount_RNA >= 350 & 
#                              percent.mito < 2.0)

#Save the seurat objects after thresholding.
#saveRDS(ss_data_decon_filtered, file = paste0( "seurat_ss_data_rethresholded", suffix, ".rds"))
```
